---
title: "Data Wrangling for Graphics"
author: "Felicia Susanto"
format: html
embed-resources: true #embed all the resources to this html file
editor: 
  markdown: 
    wrap: sentence
---

```{r libraries}
#| message: false
library(vegawidget)
library(readr)
library(dplyr)
library(vegalite)
library(lubridate)
library(vegabrite)
library(tidyr)
library(jsonlite)
```


# Exercise 1 (supply and demand)

```{r read data 1}
#| message: false
supply_demand <- read_csv("https://calvin-data304.netlify.app/data/swd-lets-practice-ex-2-3.csv")

supply_demand |>
  pivot_longer(demand:capacity, names_to = "type", values_to = "number")

supply_demand2 <- supply_demand |>
  mutate(unmet_demand = demand - capacity)
 # pivot_longer(capacity:unmet_demand, names_to = "type", values_to = "number")
  

```

```{r grouped bar chart}
vl_chart(supply_demand) |>
  vl_fold(
    c("capacity", "demand"),
    as = c("type", "value")) |>
  vl_mark_bar() |>
  vl_encode_x("date:O") |>
  vl_encode_xOffset("type:N") |>
  # vl_facet_column("type:N") |>
  vl_encode_y("value:Q") |>
  vl_encode_color("type:N", scale = list(domain = c("demand", "capacity")))
```


```{r line graph}
vl_chart(supply_demand) |>
  vl_fold(
    c("capacity", "demand"),
    as = c("type", "value")) |>
  vl_mark_line(point = TRUE) |>
  vl_encode_x("date:T") |>
  vl_encode_xOffset("type:N") |>
  vl_encode_y("value:Q") |>
  vl_encode_color("type:N", scale = list(domain = c("demand", "capacity")))
```

```{r stacked bar}
demandplot <- vl_chart(supply_demand) |>
  vl_mark_bar() |>
  vl_encode_x("date:T") |>
  vl_encode_y("demand:Q") |>
  vl_encode_color(value = "red")

capacityplot <- vl_chart(supply_demand) |>
  vl_mark_bar(width = 10) |>
  vl_encode_x("date:T") |>
  vl_encode_y("capacity:Q") |>
  vl_encode_color(value = "blue") |>
  vl_encode_opacity(value = 0.7)

vl_layer(demandplot, capacityplot)
```

```{r stacked unmet demand}
demandplot <- vl_chart(supply_demand) |>
  vl_mark_bar() |>
  vl_encode_x("date:T") |>
  vl_encode_y("demand:Q") |>
  vl_encode_color(value = "red", title = "unmet demands")

capacityplot <- vl_chart(supply_demand) |>
  vl_mark_bar() |>
  vl_encode_x("date:T") |>
  vl_encode_y("capacity:Q") |>
  vl_encode_color(value = "blue", title = "capacity") |>
  vl_encode_opacity(value = 0.7)

vl_layer(demandplot, capacityplot)
```


```{r points combined}
demandplot <- vl_chart(supply_demand) |>
  vl_mark_point() |>
  vl_encode_x("date:T") |>
  vl_encode_y("demand:Q") |>
  vl_encode_color(value = "red")

capacityplot <- vl_chart(supply_demand) |>
  vl_mark_point() |>
  vl_encode_x("date:T") |>
  vl_encode_y("capacity:Q") |>
  vl_encode_color(value = "blue")

lineplot <- vl_chart(supply_demand) |>
  vl_fold(
  c("capacity", "demand"),
  as = c("type", "value")) |>
  vl_mark_line() |>
  vl_encode_x("date:T") |>
  vl_encode_y("value:Q") |>
  vl_encode_detail("date:T")

vl_layer(demandplot, capacityplot, lineplot)
```

```{r unmet line}
vl_chart(supply_demand2) |>
  vl_mark_line(point = TRUE) |>
  vl_encode_x("date:T") |>
  vl_encode_y("unmet_demand:Q")
```


# Excercise 2 (jobs)

```{r class code}

jobs_url <- 
  "https://cdn.jsdelivr.net/npm/vega-datasets@2.8.0/data/jobs.json"


#| fig-cap: "This is a caption for the figure."
vl_chart() |>
  vl_mark_point() |>
  vl_pivot("year", groupby = list("job", "sex"), value = "perc") |>
  vl_encode_x("1850:Q") |>
  vl_scale_x(type = "symlog", constant = 0.00001) |>  # note: scale must be set quite small
  vl_encode_y("2000:Q") |>
  vl_scale_y(type = "symlog", constant = 0.00001) |>  # note: scale must be set quite small
  vl_facet_column("sex:N", title = "") |>
  vl_encode_tooltip_array(list("job", "sex", "1850", "2000")) |>
  vl_add_properties(
    title =  list( 
      text = "Percent of people working various jobs",
      subtitle = "Each dot represents the one occupation. Hover to see which occupation it is."
      )) |>
  vl_add_data_url(jobs_url) 
```

### "perc" means?
sum all values of perc in each year separated by men and women... explain what perc means

```{r read json url}
jobs_data <- fromJSON("https://cdn.jsdelivr.net/npm/vega-datasets@2.8.0/data/jobs.json")
```

```{r adding up perc}
jobs_summary <- jobs_data %>%
  group_by(year, sex) %>%
  summarise(total_perc = sum(perc, na.rm = TRUE))

glimpse(jobs_summary)
```

Looking at the data, the "perc" values add up to one for each year. This means that if we sum the percentages of men and women for the same year, the total will always be one. In other words, "perc" represents the proportion of jobs each year (where the total is 1, not 100%) for both men and women. For example, in 1850, 0.990373464 (or 99%) of workers were men, while 0.009626536 (or 1%) were women. This makes sense, as most women were likely housewives at that time.

### modifying graphic

```{r}
jobs_data_transformed <- jobs_data |>
  pivot_wider(names_from = sex, values_from = perc, id_cols = c(job, year))

# View result
head(jobs_data_transformed)
```

```{r}
vl_chart(jobs_data_transformed) |>
  vl_mark_point() |>
  vl_filter("datum.year == 2000 | datum.year == 1990 | datum.year == 1980 | datum.year == 1970") |>
  vl_facet("year:O", title = "", columns = 2) |>
  vl_encode_x("men:Q") |>
  vl_scale_x(type = "symlog", constant = 0.00001) |>  # note: scale must be set quite small
  vl_encode_y("women:Q") |>
  vl_scale_y(type = "symlog", constant = 0.00001) |>  # note: scale must be set quite small
  vl_encode_color("job:N") |>
  vl_encode_tooltip_array(list("job", "men", "women")) |>
  vl_axis_x(format=".0%") |>
  vl_axis_y(format=".0%") |>
  vl_add_properties(
    title =  list( 
      text = "Percent of people working various jobs",
      subtitle = "Each dot represents the one occupation. Hover to see which occupation it is."
      ))
```

## percentage of men and women who worked in various jobs

```{r}
#join data because jobs_summary contain total % for men and women in each year

jobs_summary <-
  jobs_summary |>
  pivot_wider(names_from = sex, values_from = total_perc, id_cols = year) |>
  rename_with(~ paste0(.x, "_total_perc"), -year)

```

```{r}
jobs_data_transformed_join <- left_join(jobs_data_transformed, jobs_summary, by = "year") |>
  mutate(
    perc_men = men / men_total_perc,
    perc_women = women / women_total_perc
  )
```

```{r}
vl_chart(jobs_data_transformed_join) |>
  vl_mark_point() |>
  vl_filter("datum.year == 2000 | datum.year == 1990 | datum.year == 1980 | datum.year == 1970") |>
  vl_facet("year:O", title = "", columns = 2) |>
  vl_encode_x("perc_men:Q") |>
  vl_scale_x(type = "symlog", constant = 0.00001) |>  # note: scale must be set quite small
  vl_encode_y("perc_women:Q") |>
  vl_scale_y(type = "symlog", constant = 0.00001) |>  # note: scale must be set quite small
  vl_encode_color("job:N") |>
  vl_encode_tooltip_array(list("job", "perc_men", "perc_women")) |>
  vl_axis_x(format=".0%") |>
  vl_axis_y(format=".0%") |>
  vl_add_properties(
    title =  list( 
      text = "Percent of people working various jobs in proportion to their gender",
      subtitle = "Each dot represents the one occupation. Hover to see which occupation it is."
      ))
```

#left join vegabrite
```
jobs_data_transformed |>
vl_lookup(
  lookup = "year",
  from = list(
    data = list(jobs_data_transformed),
    key = "year",
    fields = list("total_perc")
  )
)
```


# Exercise 3 (price at the pump)

```{r read data 3}
#| message: false
gas_prices <- read_csv("https://calvin-data304.netlify.app/data/pump_price_for_gasoline_us_per_liter.csv")

countries <- read_csv("https://raw.githubusercontent.com/lukes/ISO-3166-Countries-with-Regional-Codes/master/all/all.csv")
```

```{r joining data}

#giving both datasets common names
countries <- countries %>% rename(country = name)

joined_countries <- gas_prices %>% inner_join(countries, by = "country")

# Identify unmatched countries
unmatched_country_gas <- gas_prices %>% anti_join(countries, by = "country")  # Countries in gas_prices but not in countries
unmatched_country_countries <- countries %>% anti_join(gas_prices, by = "country")  # Countries in countries but not in gas_prices

# View results
print("Successfully Matched Data:")
print(joined_countries)

print("Countries in gas_prices but NOT in countries:")
print(unmatched_country_gas$country)

print("Countries in countries but NOT in gas_data:")
print(unmatched_country_countries$country)

```

```{r}
country_gas_final <- joined_countries|>
    pivot_longer(cols = 
        c("1991","1992","1993","1994","1995","1996", "1997","1998","1999", "2000","2001", "2002","2003", "2004","2005","2006", "2007", "2008","2009", "2010", "2011","2012", "2013", "2014", "2015", "2016"),
        names_to = "year",
        values_to = "gas_prices",
        values_transform = list(gas_prices = as.numeric))
```

```{r}
#cleaned_data_countries <- joined_countries |> drop_na()
```


```{r}
#my xOffset doesn't seem to be working in grouping the bars together per subregion...

vl_chart(country_gas_final, width = 300) |>
  vl_mark_bar() |>
  vl_filter("datum.year == 2000 | datum.year == 2002 | datum.year == 2004 | datum.year == 2006 | datum.year == 2008 | datum.year == 2010 | datum.year == 2012 | datum.year == 2014 |datum.year == 2016") |>
  vl_encode_x("year:O", title = "Year") |>
  vl_encode_xOffset("sub-region:N") |>  # Group bars by sub-region
  vl_encode_y("gas_prices:Q", title = "Gas Prices (US $)") |>
  vl_facet("region:N", columns = 2) |>   # Facet by region
  vl_encode_color("country:N", title = "Country") |>  # Color bars by country
  vl_encode_tooltip_array(list("country", "gas_prices", "sub-region", "country-code"))
```







