---
title: "Final project Data 304"
author: "felicia susanto"
format: html
editor: visual
code-tools: true
---

```{r libraries}
#| message: false
library(vegawidget)
library(readr)
library(dplyr)
library(vegalite)
library(lubridate)
library(vegabrite)
library(tidyr)
library(jsonlite)
library(readxl)
```

# Exercise 1

## a

```{r}
#| code-fold: true 
'
{
  "$schema": "https://vega.github.io/schema/vega-lite/v5.json",
  "data": {
    "url": "https://calvin-data304.netlify.app/data/twins-genetics-wide.json"
  },
  "mark": {"type": "bar"},
  "title": "Comparing Twins Genetic Similarity According to Genealogy Services",
  "height": 300,
  "width": 300,
  "encoding": {
    "x": {"field": "id", "type": "nominal", "title": ""},
    "xOffset": {"field": "region"},
    "color": {"field": "region", "title": "Region of Ancestry"},
    "y": {
      "field": "Ancestry",
      "type": "quantitative",
      "title": "Ancestry.com Score"
    },
    "facet": {"field": "pair", "columns": 3, "title": "Pairs of Twins"}
  }
}
'|> as_vegaspec()
```

This graph was taken from the homework 4 gallery and something I like about it is that I can see all the twins compared to each other in one graph. I do have to slide around a bit, but if we make the graphs smaller then they can fit into a smaller screen without needing to scroll as much. I also like the use of color which distinguishes each region so we know how each region differs for each set of twins. It also has labels on the x axis that shows individual people (so twin a and twin b for each pair of twins). This graph also allows me to compare each region, for each twin easily because bar graphs are one of the easier graphs to interpret and this graph shows all the bars at once.

## b

```{r}
#| code-fold: true 
'
{
  "$schema": "https://vega.github.io/schema/vega-lite/v5.json",
  "title": "DNA Kits Similarities/Differences",
  "data": {
    "url": "https://calvin-data304.netlify.app/data/twins-genetics-long.json",
    "format": {"type": "json"}
  },
  "height": 400,
  "width": 450,
  "mark": "boxplot",
  "encoding": {
    "x": {"field": "kit", "type": "nominal", "title": "DNA Kit"},
    "y": {
      "field": "genetic share",
      "type": "quantitative",
      "title": "Genetic Share"
    },
    "color": {"field": "kit", "type": "nominal", "title": "DNA Kit"}
  }
}
'|> as_vegaspec()
```

This graph was taken from the homework 4 gallery, and something I don't like about it is the use of color here. I think the use of color is redundant because each boxplot is already labeled and represents different DNA kits, so the use of different colors isn't really helpful. In addition, there are only 3 boxplots, so we really don't need the help of color to distinguish the plots. It honestly isn't a big deal, and the use of color would not make any difference, but that is one thing I noticed immediately. I also don't think this graphic is good at immediately telling a story because I don't understand what the plots are trying to say. What does the genetic share (y-axis) represent here, and what does it have to do with comparing the kits? Is it the average genetic share each kit finds? Or is it the genetic share each kit has compared to the total genetic share?

## c

#### comparing twins

```{r}
genetics_wide <- read.csv("https://calvin-data304.netlify.app/data/twins-genetics-wide.csv")
genetics_long <- read.csv("https://calvin-data304.netlify.app/data/twins-genetics-long.csv")
```

```         
#| code-fold: true 
vl_chart() |>
  vl_mark_bar() |>
  vl_add_properties(title ="Comparing Twins Genetic Shares") |>
  vl_encode_x(field = "twin:N", title = "Twin", sort = c(
        "1A", "1B", "2A", "2B", "3A", "3B",
        "4A", "4B", "5A", "5B", "6A", "6B"
  )) |>
  vl_encode_y(field = "genetic share:Q", title = "Genetic Share") |>
  vl_facet(field = "region", type = "nominal", columns = 2) |>
  vl_encode_color("pair:N", title = "Pairs", scale = list(scheme = "tableau10")) |>
  vl_add_data(genetics_long)
```

```{r}
#| code-fold: true 
'
{
  "$schema": "https://vega.github.io/schema/vega-lite/v5.json",
  "data": {
    "url": "https://calvin-data304.netlify.app/data/twins-genetics-long.json"
  },
  "width": 600,
  "height": 900,
  "title": "Twins genetic shares similarities/differences",
  "facet": {"field": "region", "type": "nominal", "title": "Region"},
  "spec": {
    "mark": "bar",
    "encoding": {
      "x": {
        "field": "twin",
        "type": "nominal",
        "title": "Twin",
        "sort": [
          "1A",
          "1B",
          "2A",
          "2B",
          "3A",
          "3B",
          "4A",
          "4B",
          "5A",
          "5B",
          "6A",
          "6B"
        ]
      },
      "y": {
        "field": "genetic share",
        "type": "quantitative",
        "title": "Genetic Share"
      },
      "color": {
        "field": "pair",
        "type": "nominal",
        "title": "Pairs",
        "scale": {"scheme": "tableau10"}
      },
      "tooltip": [
        {"field": "genetic share", "type": "quantitative", "title": "Genetic Share"},
        {"field": "kit", "type": "nominal", "title": "Kit"}
      ]
    }
  }
}
' |> as_vegaspec()

```

This graph tells the story of genetic sharing among twins, allowing us to compare these shares between each twin pair. The tooltip provides valuable insight by displaying the kit used and the corresponding genetic share it contributes to each twin pair. It also shows the overall genetic shares between the twins. The stacked bar chart is helpful in conveying this information, and the use of distinct colors and grouped bars facilitates easy comparison of twins in each pair. Moreover, the facet by region feature enables us to observe the regional genetic contributions, which, ideally, should be similar for each twin individual when compared to their twin.

#### comparing kits

```{r}
#| code-fold: true 
vl_chart(width = 300) |> 
  vl_mark_bar() |> 
  vl_encode_x("kit:N") |>
  vl_encode_y("genetic.share:Q", title = "genetic share") |>
  vl_encode_color("pair:N") |> 
  vl_encode_xOffset("twin:O") |> 
  vl_facet("pair:N") |>
  vl_encode_tooltip_array(list("twin", "genetic.share", "region", "kit")) |>
  vl_add_properties(title = "comparing 3 kits across twins")|>
  vl_add_data(genetics_long)
```

This graph tells a story about how each kit is compared to one another in their genetic share. I faceted by twin so it's easier to see exactly how each kit is different for each twin side by side. The tooltip also helps give extra information like more detailed genetic share for each part of the bar graph. For some kits, the genetic share is seen to be over the value 1 (twin 3A, 23andMe) which I don't think should happen because the total genetic share should be 1. From this graph, we can tell the accuracy and compare differences across kits.

## d

story under each graph.

# Exercise 2

#### email marketing challenge

```{r}
challenge_data <- fromJSON("https://calvin-data304.netlify.app/data/swd-lets-practice-ex-2-13.json")
```

This data shows the completion rates for an email marketing campaign and also the response rate. The problem is that the response rate is low and the completion rate is high, but the completion rate comes from the total people who responded, so it shouldn't be more than the response rate. There should be a new variable that calculates the completion rate with respect to the response rate. So how many people actually completed the survey, under the amount of people that actually responded.

```{r}
#| code-fold: true
challenge_data <- challenge_data |>
  separate(Date, into = c("quarter", "year"), sep = "-") |>
  mutate(
    quarter = parse_number(quarter),
    year = parse_number(year),
    date_str = paste0(year, "-", 3 * quarter - 2, "-01" ),
    date = ymd(date_str))
  
challenge_data <- challenge_data |> mutate(TrueCompletionRate = `Completion Rate` * `Response Rate`)
```

```{r}
#| code-fold: true

responseplot <- vl_chart(challenge_data) |>
  vl_mark_bar() |>
  vl_encode_x("date:T") |>
  vl_encode_y("Response Rate:Q") |>
  vl_encode_color(value = "blue")|>
  vl_encode_tooltip_array(list("Response Rate"))

completionplot <- vl_chart(challenge_data) |>
  vl_mark_line() |>
  vl_encode_x("date:T") |>
  vl_encode_y("TrueCompletionRate:Q") |>
  vl_encode_color(value = "red")

completionpoint <- vl_chart(challenge_data) |>
  vl_mark_point() |>
  vl_encode_x("date:T") |>
  vl_encode_y("TrueCompletionRate:Q") |>
  vl_encode_color(value = "red")|>
  vl_encode_tooltip_array(list("TrueCompletionRate", "Completion Rate", "Response Rate"))

vl_layer(responseplot, completionplot, completionpoint) |>
  vl_add_properties(title = "Completion Rates in Proportion to Response Rate")
```

# Exercise 3

## a

```         
##Idk why this isn't working, but i can access the file through pulling it through my laptop.
Tanzania_Data <- read_csv("https://github.com/Felicia-susanto/data304/raw/refs/heads/main/data/Tazania%20Data.xlsx")
```

```{R}
#just so the code can run from other devices
tanzania <- data.frame(
  Survey_Year = c(1991, 1992, 1996, 1999, 2004, 2005, 2010, 2015, 2016),
  Fertility_Rate = c(6.2, 6.2, 5.8, 5.6, 5.7, 5.7, 5.4, 5.2, 5.2),
  Contraception_Use = c(0.059, 0.059, 0.117, 0.156, 0.176, 0.176, 0.236, 0.384, 0.384),
  Unmet_FP = c(0.278, 0.278, 0.26, 0.223, 0.243, 0.243, 0.223, 0.221, 0.221)
)
```

For surveys that were conducted in 2 years, I duplicated their values and put both years in with the same value. I also changed all the percentages to be proportions for now.

## b

```         
FR <- vl_chart(Tazania_Data) |>
  vl_mark_line() |>
  vl_encode_x("Survey_Year:O") |>
  vl_encode_y("Fertility_Rate:Q") |>
  vl_encode_color(value = "blue")|>
  vl_encode_tooltip_array(list("Fertility_Rate", "Unmet_FP", "Contraception_Use"))

Contraception <- vl_chart(Tazania_Data) |>
  vl_mark_line() |>
  vl_encode_x("Survey_Year:O") |>
  vl_encode_y("Contraception_Use:Q") |>
  vl_encode_color(value = "red")|>
  vl_encode_tooltip_array(list("Fertility_Rate", "Unmet_FP", "Contraception_Use"))

FP <- vl_chart(Tazania_Data) |>
  vl_mark_line() |>
  vl_encode_x("Survey_Year:O") |>
  vl_encode_y("Unmet_FP:Q") |>
  vl_encode_color(value = "orange")|>
  vl_encode_tooltip_array(list("Fertility_Rate", "Unmet_FP", "Contraception_Use"))

FR2 <- vl_chart(Tazania_Data) |>
  vl_mark_point() |>
  vl_encode_x("Survey_Year:O") |>
  vl_encode_y("Fertility_Rate:Q") |>
  vl_encode_color(value = "blue")|>
  vl_encode_tooltip_array(list("Fertility_Rate", "Unmet_FP", "Contraception_Use"))

Contraception2 <- vl_chart(Tazania_Data) |>
  vl_mark_point() |>
  vl_encode_x("Survey_Year:O") |>
  vl_encode_y("Contraception_Use:Q") |>
  vl_encode_color(value = "red")|>
  vl_encode_tooltip_array(list("Fertility_Rate", "Unmet_FP", "Contraception_Use"))

FP2 <- vl_chart(Tazania_Data) |>
  vl_mark_point() |>
  vl_encode_x("Survey_Year:O") |>
  vl_encode_y("Unmet_FP:Q") |>
  vl_encode_color(value = "orange")|>
  vl_encode_tooltip_array(list("Fertility_Rate", "Unmet_FP", "Contraception_Use"))

vl_layer(FR, Contraception, FP, FR2, Contraception2, FP2) |>
  vl_add_properties(title = "Fertility Rate, Contraception, and Family Planning in Tanzania")
```

```{r}
#| code-fold: true
#This gives a legend for color
Tanzania_Data_long <- tanzania|>
  pivot_longer(cols = c(Fertility_Rate, Contraception_Use, Unmet_FP), 
               names_to = "Category", 
               values_to = "Value")
vl_chart(Tanzania_Data_long) |>
  vl_mark_line() |>
  vl_encode_x("Survey_Year:O") |>
  vl_encode_y("Value:Q", title = "Rate") |>
  vl_encode_color(field = "Category:N", 
                  title = "Indicator", 
                  scale = list(domain = c("Fertility_Rate", "Unmet_FP", "Contraception_Use"), 
                               range = c("blue", "orange", "red"))) |>
  vl_encode_tooltip_array(list("Survey_Year", "Category", "Value")) |>
  vl_add_properties(title = "Women's Fertility Rate, Contraception, and Family Planning in Tanzania")

```

```{r}
#this uses repeat
vl_chart(Tanzania_Data_long) |>
  vl_mark_line() |>
  vl_encode_x("Survey_Year:O") |>
  vl_encode_y("Value:Q") |>
  vl_encode_color("Category:N") |>
  vl_repeat_col("Category:N") |>
  vl_encode_facet("Category:N", title = "") |>
  vl_add_properties(title = "Tanzania Women's fertility Over Time")
```

## c

This graph tells a story about women's fertility rates, contraception use, and unmet family planning needs in Tanzania over time.

Fertility Rate (Blue Line): The fertility rate has declined steadily from around 6 children per woman in the early 1990s to around 5 in the 2010s. This suggests improvements in reproductive health education, economic factors, and possibly access to family planning services.

Contraception Use (Red Line): The use of contraception has gradually increased over time, though it looks like it remains relatively low compared to fertility rates, we must remember that it is a proportion, so it only goes up to 1. This could indicate increased awareness and availability of contraceptive methods.

Unmet Family Planning Needs (Orange Line): The unmet need for family planning appears relatively low and stable over the years. If this line is declining, it suggests that more women are gaining access to the contraception they need. However, if it remains constant or rises, it indicates persistent barriers.

The declining fertility rate is possibly influenced by increasing contraception use and improved family planning services. However, unmet family planning is still relatively stable, meaning there might still be cultural, economic, or logistical barriers limiting access to some families.

# Exercise 4

```{r}
penguins <- fromJSON("https://raw.githubusercontent.com/vega/vega-datasets/refs/heads/main/data/penguins.json")
```

```{r}
#| code-fold: true
penguinsplot <- vl_chart(penguins) |>
  vl_add_properties(title = "Penguin Measurements Interactive Plot") |>
  vl_mark_point() |>
  vl_encode_x("Beak Length (mm):Q", title = "Beak Length (mm)", scale = list(domain = list(20, 90))) |>
  vl_encode_y("Flipper Length (mm):Q", title = "Flipper Length (mm)", scale = list(domain = list(170, 240))) |>
  vl_encode_size("Body Mass (g):Q", title = "Body Mass (g)") |>
  vl_encode_color(
    "Species:N",
    title = "Species",
    scale = list(
      domain = list("Adelie", "Gentoo", "Chinstrap"),
      range = list("steelblue", "goldenrod", "tomato")
    )
  ) |>
  vl_encode_tooltip_array(list(
    "Beak Length (mm)",
    "Flipper Length (mm)",
    "Body Mass (g)",
    "Island",
    "Species"
  ))

penguinsplot |>
  vl_add_parameter("width", value = 200) |>
  vl_add_parameter("height", value = 200) |>
  vl_bind_range_input("width", min = 100, max = 400, step = 1) |>
  vl_bind_range_input("height", min = 100, max = 400, step = 1) |>
  vl_add_properties(
    width = list(expr = "width"),
    height = list(expr = "height")
  ) |>
  vl_add_parameter("Island", value = "All") |> 
  vl_bind_select_input(
    parameter_name = "Island",
    name = "Island: ",
    options = list("All", "Torgersen", "Dream", "Biscoe")
  ) |>
  vl_filter("Island == 'All' || datum.Island == Island")|>
  vl_add_point_selection(
    "species_param",
    bind = "legend",
    encodings = list("color")
  ) |>
  vl_add_point_selection(
    "size_param",
    bind = "legend",
    encodings = list("size")
  ) |>
  vl_encode_opacity(value = 0.2) |>
  vl_condition_opacity(param = "species_param", value = 1) |>
  vl_condition_opacity(param = "size_param", value = 1)

```

I chose a scatterplot because it best represents the data I’m trying to encode. Both beak length and flipper length are quantitative variables, so a scatterplot is the most effective way to display their relationship. I made sure that the scale for both axes remains consistent even when zooming in, ensuring that the jump between each point is the same. Since both variables are measured in millimeters, I kept the scales aligned for accurate visual comparison.

I chose not to start the axes at zero, and that’s okay because this is a scatterplot, not a bar chart, where starting at zero is necessary for comparing proportions. The key was maintaining the same scale for both axes so the comparison remains meaningful.

I added interactive legends and tooltips to make the graph more engaging and informative. Just by hovering over a few dots, we can quickly access a lot of information, which is helpful since many of the points are clustered in one area. The interactive legend also allows for adjusting the visibility of different species and body mass sizes. By adding opacity adjustments, the graph becomes more readable—areas with higher point density appear darker, making it easier to see where most data points are concentrated.

I considered other visualizations, such as using facet plots by island or a heatmap, but I felt that a scatterplot was the clearest way to show the relationships between these variables. The facets would have separated the data too much, making it harder to compare species across islands, and a heatmap wouldn't have captured the individual data points as effectively.

# Exercise 5

-   an encoding channel other than x or y: used color for most graphs (comparing twins, kits, challenge, tanzania )

-   layers: Used in exercise 2 to graph completion rates and response rates

-   facets: used in comparing kits and comparing twins

-   concatenation or repeat: tanzania health repeats

-   non-default settings for a channel’s scale or guide: used tableau10 for color in comparing twins, sorted the twin field as well under vl_encode_x, mutate date for challenge.

-   tooltips: used in most of my graphics: exercise 2, exercise 3

-   another kind of interaction (panning/zooming, brushing, sliders, etc.): used dropdown, sliders, and interactive legends for exercise 4

# Exercise 6

#### a

In my masterpiece, I learned how to add the option "all" in a dropdown box that lets us use all of the different options in a variable all at once. It's like showing how the graph would be without the filtering while still being able to choose specific kinds of variable types as well

#### b

In wilke chapter 6, he mentions that the use of bar charts are bad for axes that don't start at 0, because of that, I used a scatter plot like he suggested instead. I felt like I had to zoom into the penguins graph because most of the points were concentrated in the upper right corner of the graph, so if I zoom the graph in, then it would make it easier to see individual points and the spread of the points.

He also mentions the importance of keeping the scales of the x and y axis the same (especially in my case where I am comparing beak and flipper sizes and both are in mm) so that it would not be misleading, so I implemented both of these in my graphs.
