---
title: "Maps, Pies, and Other Graphics"
author: "Felicia Susanto"
format: html
embed-resources: true #embed all the resources to this html file
editor: 
  markdown: 
    wrap: sentence
---

```{r libraries}
#| message: false
library(vegawidget)
library(readr)
library(dplyr)
library(lubridate)
library(vegabrite)
library(tidyr)
```

# Exercise 1

## a
Read chapter 4 of of Wilke (2019). List two or three things you learned about using color from this chapter.

* I learned that color is a great tool for helping readers distinguish between items or groups that do not have an inherent order. The colors used for this purpose should be clearly distinct from one another while remaining equivalent. This means that no color should stand out relative to the others or create the impression of an order. For example, when mapping colors to U.S. states, each state is distinct but does not follow a particular order.

* For data with values that have a natural order, such as temperature, weight, or income, we can use a sequential color scale. These color scales contain colors arranged in a clear progression, indicating which values are larger or smaller. For example, dark to light blue can represent height, while dark to light red can represent weight. However, the hues do not have to be a single color; a rainbow gradient can also be used. When using multiple hues, we must be cautious because unnatural transitions, such as dark yellow to light blue, may appear visually confusing and ineffective for a sequential scale.

* A diverging color scale is useful for datasets containing both positive and negative numbers, as it effectively highlights deviations from zero. This type of scale consists of two sequential color scales stitched together at a common midpoint, usually a light color. The progression from the midpoint to the darker colors on either end should be balanced to maintain visual clarity and consistency.

* Colors can also be used for highlighting, but it is important to ensure that the baseline colors do not compete for attention. Instead, they should support and complement the accent color.


## b
Read chapter 5 of of Wilke (2019). Chapter 5 list a wide range of graphic types. What is the organization scheme Wilke uses to group them? What is the most interesting thing you learned from this chapter?

Wilke organizes the various graphic types in Chapter 5 based on the type of information they want their data to visualize. The chapter groups visualizations into the following categories:

* Amounts – Visualizations that represent numerical values for a set of categories (e.g., bar charts, dot plots, heatmaps).
* Distributions – Charts that show how data is spread out, such as histograms, density plots, and boxplots.
* Proportions – Visualizations that compare parts to a whole, like pie charts, stacked bars, and treemaps.
* x–y Relationships – Charts that depict relationships between two or more quantitative variables, including scatterplots, bubble charts, and line graphs.
* Geospatial Data – Maps and other spatial visualizations, such as choropleths and cartograms.
* Uncertainty – Graphics that illustrate confidence intervals, distributions, and error margins, such as error bars, confidence bands, and quantile dot plots.


The most interesting insight from this chapter is the discussion on uncertainty visualization. Traditional error bars are common, but Wilke highlights alternatives like graded error bars, confidence strips, and quantile dot plots, which can provide a clearer or more intuitive representation of uncertainty. The idea of using "eyes" and "half-eyes" to combine error bars with distribution visualizations is particularly intriguing because it merges both precise confidence ranges and overall uncertainty distributions.


## c
By now, you should be able to make quite a few of the graphics shown in chapter 5. Are there some you would not know how to make (yet) based on what we have done in class? List several that you don’t know how to make – choose the ones you would be most interested in learning how to make and say why you are interested.

List of graphs I don't know how to make:

* Box plots: I am interested in learning how to make a box plot because it is a very informative graphic that displays the median, minimum, maximum, and quantiles.

* Ridgeline plots: This is an interesting graph, and I assume that creating it would require the use of layers, but I am not sure.

* Stacked density: I would like to learn how to graph proportion changes along a continuous variable.

* Eyes: This is a great graph that combines error bars with their distributions. It would be interesting to learn how to create.


## d
See if you can figure out how to make at least one of the types of graphic listed in part c.  You may need to do some experimenting or to consult the documentation or example galleries (for Vega-Lite, vegabrite, altiar, Altair, etc.).

I found how to make a boxplot in the vegabrite documentation. Here is the code:

```{r}
vl_chart(data = 'https://vega.github.io/editor/data/penguins.json') %>%
  vl_mark_boxplot() %>% # the mark I needed to use
  vl_encode_x('Species:N') %>%
  vl_encode_y('Body Mass (g)', type = 'quantitative') %>%
  vl_scale_y(zero = FALSE) %>%
  vl_encode_color('Species:N', legend = NA)
```

```{r}
#my own boxplot

df <- data.frame(
  date = c("Q1-2017", "Q2-2017", "Q3-2017", "Q4-2017", "Q1-2018", "Q2-2018", "Q3-2018", "Q4-2018", "Q1-2019"),
  completion_rate = c(0.91, 0.93, 0.91, 0.89, 0.84, 0.88, 0.91, 0.87, 0.83),
  response_rate = c(0.023, 0.018, 0.028, 0.023, 0.034, 0.027, 0.026, 0.039, 0.028)
)

df_pivot <- df |>
  separate(date, into = c("Quarter", "Year"), sep = "-") |>
  pivot_longer(
    cols = c(completion_rate, response_rate),
    names_to = "rate_type",
    values_to = "value"
  ) |>
  mutate(rate_type = as.factor(rate_type))
```

```
vl_chart(df_pivot) |>
  vl_mark_boxplot() |>
  vl_encode_x("Year:N") |>
  vl_encode_y("value:Q") |>
  vl_encode_color("Year:N") |>
  vl_facet(field = "rate_type", columns = 2) |>
  vl_scale_y(zero = FALSE) |>
  vl_encode_tooltip(c("Year:N", "value:Q"))

#It's giving me this error: Error in !hasName(obj, "type") && hasName(obj, "field") && grepl(":[N,O,Q,T]$",  : 
  #'length = 2' in coercion to 'logical(1)'

```


## e
Chapters 6 - 16 are orgainized by graphic types. You have already read some of this material. Pick one chapter that is related to your graphic in part d and give some examples of how what you learned in that chapter affected how you created your graphic.

From the box plot, we can see that the Gentoo species has the highest body mass on average, while the Chinstrap species has a wider range of body mass values with several outliers. Interestingly, despite this variability, the Chinstrap and Adélie species share the same median body mass. The use of color in the plot helps differentiate the three species, making it easier to compare their distributions at a glance.

In Chapter 97, which covers box plots and violin plots, I learned about the importance of visualizing distributions using quartiles and outliers. This influenced my decision to use a box plot, as it clearly shows the median, interquartile range, and outliers, which are essential for understanding the variability in body mass among the penguin species. Additionally, the chapter emphasized the importance of scaling the y-axis appropriately to avoid misleading representations, which is why I set vl_scale_y(zero = FALSE) to ensure the plot accurately reflects variations in body mass without unnecessary distortion and allowing the boxplots to fill most of the graph without too much empty space in values under 3.000g.


# Exercise 2

## a
Create side-by-side pie charts with good labeling and appropriate choice of colors. Some more things to consider: Should you present using counts or percentages? How will you deal with the nonresponders?

```{r}
pie_data <- read_csv("https://calvin-data304.netlify.app/data/likert-survey.csv")
pie_data <- pie_data |>
  group_by(year)|>
  mutate(proportion = count / sum(count))
```

```{r}
pie_layer <- vl_chart(data = pie_data) |>
  vl_encode_theta(field = "proportion:Q") |>
  vl_encode_color(field = "response:N") |>
  vl_facet(field = "year:O", title = "")|>
  vl_encode_tooltip("count") |>
  vl_mark_arc()

pie_layer

# text_layer <- vl_chart(data = pie_data) |>
#   vl_mark_text(radius = 105) |>
#   vl_encode_text('response:N')
# 
# text2_layer <- vl_chart(data = pie_data) |>
#   vl_mark_text(radius = 40) |>
#   vl_encode_text('proportion:Q')
#  
# vl_layer(pie_layer, text_layer, text2_layer)
```

## b 
Create some other kind of graphic to visualize the same data.

```{r}
vl_chart(pie_data)|>
  vl_mark_bar() |>
  vl_encode_x("response:N", sort = list(op = "values", values = c("strongly disagree", "disagree", "neither agree or disagree", "agree", "strongly agree", "no response"))) |>
  vl_encode_y("count:Q") |>
  vl_facet("year:O")|>
  vl_encode_tooltip(field = "count", "proportion")
  
```

```{r}
vl_chart(pie_data)|>
  vl_mark_bar() |>
  vl_encode_x("response:N", sort = list(op = "values", values = c("strongly disagree", "disagree", "neither agree or disagree", "agree", "strongly agree", "no response"))) |>
  vl_encode_y("proportion:Q") |>
  vl_facet("year:O")|>
  vl_encode_tooltip(field = "proportion")
```

  

## c
Comment on the pros and cons of your two (or more) graphics. Which would you recommend (for what purposes)?

When comparing counts and proportions, bar charts offer a clear and effective visualization. Additionally, bar charts are easier to interpret at a glance, as they align with our natural ability to compare lengths and heights more accurately than angles or areas.

On the other hand, pie charts, are used to represent proportions, can be space-inefficient when visualizing multiple sets of proportions or comparing changes across conditions. They often obscure relationships because human perception struggles with accurately judging the relative sizes of pie slices, especially when there are many categories with small differences. While pie charts can be useful for showing simple part-to-whole relationships, they become less effective as the number of segments increases.

For comparing proportions across multiple conditions, grouped bar charts are generally a better option, provided the number of conditions remains manageable. If the dataset includes a large number of conditions, stacked bar charts may be a more practical alternative, as they conserve space while still allowing for comparisons within and across categories.

For most comparative tasks, a bar chart—either grouped or stacked—is the preferred choice due to its clarity, efficiency, and ease of interpretation. Pie charts may still be useful in cases where the primary goal is to emphasize a single proportion out of a whole rather than making precise comparisons across conditions. However, for datasets requiring multi-condition comparisons or trend analysis, bar charts remain the superior option.


# Exercise 3
(Create a choropleth map) Create a choropleth map of the United States (by county or by state) or the worlds (by country) using any data you like. See the maps slides for some possibilities, but feel free to get creative with your data.


```{r}
library(altair)
```


















































